---
title: "Theory: Dominance Effects, Part 2"
output: html_notebook
---

This document will demonstrate the quantitative genetic of heterosis and show heterosis in an example maize simulation.

## Quantitative genetics of heterosis

The below simulation will show different types of heterosis presented in the Labroo et. al. figure.

```{r}
library(AlphaSimR)

# Adding a split in coalescent simulation to model 2 populations that
# diverge 100 generations ago
founderPop = runMacs(nInd=200, nChr=10, segSites=1000, 
                     split=100)

SP = SimParam$
  new(founderPop[1:100])$
  addTraitAD(1000, meanDD=0.2, varDD=0.1)

# Splitting the two populations in founderPop into two separate populations
A = newPop(founderPop[1:100])
B = newPop(founderPop[101:200])

# Creating fully inbred versions of the two populations
A_inbred = makeDH(A)
B_inbred = makeDH(B)

# Creating an F1 hybrids between populations A and B
F1 = hybridCross(A, B)

# Creating F1 hybrids using the inbred populations
F1_inbred = hybridCross(A_inbred, B_inbred)

# Creating an F2 population by random mating the F1s
F2 = randCross(F1, nCrosses=10000)

# Examine the means of the initial populations
meanG(A)
meanG(B)

# Examing the means of the inbred populations
meanG(A_inbred)
meanG(B_inbred)

# Examine the means of the F1 populations
# They should be approximately equivalent
meanG(F1)
meanG(F1_inbred)

# Examining mean of F2
meanG(F2)

# Panmictic Midparent Heterosis
meanG(F1) - (meanG(A)+meanG(B))/2

# Inbred Midparent Heterosis
meanG(F1) - (meanG(A_inbred)+meanG(B_inbred))/2

# Baseline Heterosis
(meanG(A)+meanG(B))/2 - (meanG(A_inbred)+meanG(B_inbred))/2

# F2 Heterosis (about half of panmictic midparent heterosis)
meanG(F2) - (meanG(A)+meanG(B))/2


## Extra not in heterosis figure

# Create F2 by selfing F1
F2_self = self(F1)

# Selfing removes half of inbred midparent heterosis
meanG(F2_self) - (meanG(A_inbred)+meanG(B_inbred))/2

```

## Maize example

The below script simulates a maize population using parameters determined by after tuning. It will also show the correlation between per se and testcross performance.

```{r}
founderPop = runMacs(nInd=200, nChr=10, segSites=300, 
                     split=100, species="MAIZE", inbred=TRUE)

SP = SimParam$
  new(founderPop[1:100])$
  addTraitAD(300, 
             mean=70, # Bushels per acre 
             var=20, # Inbred variance
             meanDD=0.92, 
             varDD=0.3)$
  setVarE(H2=1)

# Create inbred populations
A = newPop(founderPop[1:100])
B = newPop(founderPop[101:200])

# Create all possible hybrids
F1 = hybridCross(A, B)

# Examine means of populations
meanG(A)
meanG(B)
meanG(F1)

# Examine variances
varG(A)
varG(B)
varG(F1)

# Calculating GCA using built-in functions and measuring
# GCA - per se correlation
GCA = calcGCA(F1, use="gv")

head(GCA$GCAf)

cor(GCA$GCAf[,2], gv(A))

cor(GCA$GCAm[,2], gv(B))

# Testcross with 1 tester

A = setPhenoGCA(A, testers=B[1])
B = setPhenoGCA(B, tester=A[1])

cor(GCA$GCAf[,2], pheno(A))
cor(GCA$GCAm[,2], pheno(B))

# Testcross with 3 testers

A = setPhenoGCA(A, testers=B[1:3])
B = setPhenoGCA(B, tester=A[1:3])

cor(GCA$GCAf[,2], pheno(A))
cor(GCA$GCAm[,2], pheno(B))
```
