---
title: "Theory: Dominance Effects, Part 3"
output: html_notebook
---

The document will demonstrate the use of a new function for defining traits with dominance effects by supplying inbreeding depression and dominance variance. The function uses R's `optim` function to attempt to find values for mean and variance of dominance degrees that satisfy the user's request subject to limits on the range of these values. If a solution is not found, the closest solution is return. This function has not been thoroughly tested and is still somewhat experimental.

```{r}
library(AlphaSimR)

founderPop = runMacs(nInd=1000, nChr=10, segSites=1000)

SP = SimParam$new(founderPop)

# New function for specifying a trait with dominance effects
SP$altAddTraitAD(nQtlPerChr = 1000, 
                 mean = 0, 
                 varA = 1, 
                 varD = 0.2, 
                 inbrDepr = 2)


pop = newPop(founderPop)

gp = genParam(pop)

gp$varA

gp$varD

# Inbreeding depression (F=current - F=1)
mean(gp$gv_d) - gp$mu 

# Inbreeding depression (F=0 - F=1)
# This is the definition of inbreeding depression used in the function above
mean(gp$gv_d) - gp$mu + gp$mu_HW

# Manually compute inbreeding depression (F=0 - F=1)
p = colMeans( pullQtlGeno(pop) ) / 2
q = 1-p
d = SP$traits[[1]]@domEff

sum(2*p*q*d)

```

The below script shows how changing the number of QTL may be necessary to achieve a desired scenario.

```{r}
founderPop = runMacs(nInd=1000, nChr=10, segSites=1000)

SP = SimParam$new(founderPop)

SP$altAddTraitAD(nQtlPerChr = 100, 
                 mean = 0, 
                 varA = 1, 
                 varD = 0.3, 
                 inbrDepr = 10, 
                 limMeanDD = c(0, 0.6), 
                 limVarDD = c(0, 0.5))

SP$altAddTraitAD(nQtlPerChr = 1000, 
                 mean = 0, 
                 varA = 1, 
                 varD = 0.3, 
                 inbrDepr = 10, 
                 limMeanDD = c(0, 0.6), 
                 limVarDD = c(0, 0.5))
```
