---
title: "Theory: Polyploids"
output: html_notebook
---

This document will show how to simulate polyploids in AlphaSimR and demonstrate several behaviors that are unique to polyploids.

## Simulating polyploids

This script shows how to create polyploids with `runMacs` and how to set the probability of quadrivalent pairing.

```{r}
library(AlphaSimR)

# Changing ploidy argument in runMacs
founderPop = runMacs(nInd=100, nChr=10, segSites=1000, 
                     ploidy=4)

founderPop

SP = SimParam$
  new(founderPop)$
  addTraitAD(1000, meanDD=0.2, varDD=0.1)$
  setVarE(H2=1)

# Variable for probability of quadrivalent pairing
SP$quadProb

pop = newPop(founderPop)

head(
  pullMarkerGeno(pop, 
                 markers=c("1_1", "2_1", "3_1", 
                           "4_1", "5_1", "6_1")),
  n=8
  )

head(
  pullMarkerHaplo(pop, 
                  markers=c("1_1", "2_1", "3_1", 
                            "4_1", "5_1", "6_1")),
  n=8
)
```

## Double reductions

Double reductions result in two copies of a chromosome segment being passed to a gamete. This behavior can only occur in multivalent paring. This script will indirectly show the presence of double reductions by tracking inbreeding depression.

```{r}
# Create variable to track population means
muA = muB = numeric(31)
A = B = pop

# Bivalent pairing only
SP$quadProb = 0

muA[1] = meanG(A)

for(i in 2:31){
  A = self(A)
  muA[i] = meanG(A)
}

# Quadrivalent pairing only
SP$quadProb = 1

muB[1] = meanG(B)

for(i in 2:31){
  B = self(B)
  muB[i] = meanG(B)
}

plot(0:30, muA, type="l", 
     ylab="Mean", xlab="Selfing Generation")
lines(0:30, muB, col="red")


```

## Contribution of digenic dominance to response to selection

The digenic dominance in polyploids behaves similarly to additive-by-additive genetic variance with respect to response to selection. This will be demonstrated in the below script.

```{r}
founderPop = quickHaplo(nInd=1000, nChr=10, 
                        segSites=1000, ploidy=4)

SP = SimParam$
  new(founderPop)$
  addTraitAD(1000, varDD=2)$
  setVarE(H2=1)

pop = newPop(founderPop)

varA(pop)
varD(pop)

pop2 = selectCross(pop, nInd=100, nCrosses=1000)

# Observed response to selection
meanG(pop2) - meanG(pop)

# Expected response without digenic dominance
selInt(p=0.1) * varA(pop) / sqrt(varP(pop))

# Expected response with digenic dominance in an unselected population
selInt(p=0.1) * (varA(pop) + (1/3)*varD(pop)) / sqrt(varP(pop))

# Repeat measurement after multiple rounds of selecton
for(i in 1:10){
  pop = selectCross(pop, nInd=100, nCrosses=1000)
}

pop2 = selectCross(pop, nInd=100, nCrosses=1000)

# Observed response to selection
meanG(pop2) - meanG(pop)

# Expected response without digenic dominance
selInt(p=0.1) * varA(pop) / sqrt(varP(pop))

# Expected response with digenic dominance in an unselected population
selInt(p=0.1) * (varA(pop) + (1/3)*varD(pop)) / sqrt(varP(pop))

```

## Progressive heterosis

The below script will demonstrate the occurrence of progressive heterosis in tetraploids and its absence in diploids. This script comes from the AlphaSimR_Examples GitHub repository. It uses 100 replications, because progressive heterosis is not as reliably present as many of the other properties that have been presented.

```{r}
het2sc = het4sc = het2dc = het4dc = numeric(100)
for(REP in 1:100){
  founderPop = quickHaplo(nInd=400, nChr=4, segSites=100, ploidy=4)
  
  SP = SimParam$
    new(founderPop)$
    addTraitAD(100, meanDD=0.6, varDD=0.2)$
    setVarE(h2=0.3)
  
  # Create 4 tetraploid populations
  tetraploid = vector("list", 4)
  tetraploid[[1]] = newPop(founderPop[1:100])
  tetraploid[[2]] = newPop(founderPop[101:200])
  tetraploid[[3]] = newPop(founderPop[201:300])
  tetraploid[[4]] = newPop(founderPop[301:400])
  
  # Create diploid populations from tetraploid populations
  diploid = vector("list", 4)
  for(i in 1:4){
    diploid[[i]] = reduceGenome(tetraploid[[i]])
  }
  
  # Perform 5 generations of recurrent selection
  for(gen in 1:5){ 
    for(i in 1:4){
      tetraploid[[i]] = selectCross(tetraploid[[i]], nInd=10, nCrosses=100)
      diploid[[i]] = selectCross(diploid[[i]], nInd=10, nCrosses=100)
    }
  }
  
  # Create single cross populations (1/2 and 3/4)
  tetraSC = list(randCross2(tetraploid[[1]], tetraploid[[2]], nCrosses=100),
                 randCross2(tetraploid[[3]], tetraploid[[4]], nCrosses=100))
  diploidSC = list(randCross2(diploid[[1]], diploid[[2]], nCrosses=100),
                   randCross2(diploid[[3]], diploid[[4]], nCrosses=100))
  
  # Create double cross populations
  tetraDouble = randCross2(tetraSC[[1]], tetraSC[[2]], nCrosses=1000)
  diploidDouble = randCross2(diploidSC[[1]], diploidSC[[2]], nCrosses=1000)
  
  # Calculate mean of all base populations
  meanTetra = mean(unlist(lapply(tetraploid,meanG)))
  meanDiploid = mean(unlist(lapply(diploid,meanG)))
  
  # Calculate mean of single cross populations
  meanTetraSC = mean(unlist(lapply(tetraSC,meanG)))
  meanDiploidSC = mean(unlist(lapply(diploidSC,meanG)))
  
  # Measure single cross heterosis
  het4sc[REP] = meanTetraSC - meanTetra
  het2sc[REP] = meanDiploidSC - meanDiploid
  
  # Measure double cross heterois
  het4dc[REP] = meanG(tetraDouble) - meanTetra 
  het2dc[REP] = meanG(diploidDouble) - meanDiploid 
}
boxplot(het2dc,het2sc,het4dc,het4sc, 
        names=c("Diploid D", "Diploid S",
                "Tetraploid D", "Tetraploid S"),
        las=2,
        main="Single (S) vs Double (D) Cross Means")
boxplot(het2dc-het2sc,het4dc-het4sc, 
        names=c("Diploid", "Tetraploid"),
        las=2,
        main="Progressive Heterosis")

```

## Other polyploid behaviors

```{r}
founderPop = quickHaplo(nInd=100, nChr=10, 
                        segSites=1000)

SP = SimParam$
  new(founderPop)$
  addTraitAD(1000, meanDD=0.6)

diploid = newPop(founderPop)

# Create tetraploids from diploids
tetraploid = doubleGenome(diploid)
tetraploid2 = randCross(tetraploid, 1000)

# Polyploids produced by doubling are partially inbred (not HWE)
meanG(tetraploid2) - meanG(tetraploid)

# You can cross different ploidy levels
triploid = randCross2(diploid, tetraploid, 1000)

# Odd ploidy levels are dead ends
```
